Supplementary material

### 2018-03-01 Manuscript/ms.tex
Retooling the discussionn

As discussed above, the race between mosquito mortality and incubation period introduces a critical bottleneck and non-linearity into the transmission process and the effect of mosquito abundance on human disease.
Given the central role that this process plays in driving dengue dynamics, we model fluctuations in the mosquito mortality rate using a flexible, stochastically forced harmonic oscillator.
The fact that this is the only source of stochasticity in the transmission process, however,  means that the estimated mosquito mortality time series could be soaking up other sources of stochasticity or model misspecification.

Our implementation of the mosquito mortality process is conceptually similar to the empirical forcing functions developed by Hooker and Ellner \cite{Hooker2015}.
In that framework, Hooker and Ellner \cite{Hooker2015} estimated nonparametric functions in time that modify either the differential equations or specific model parameters to provide a good fit to the data.
These forcing functions serve as residuals on the time derivatives, and can be more readily interpreted as indicators of lack-of-fit than residuals on the state variables.
We do not employ the same explicit goodness-of-fit testing framework as \cite{Hooker2015}, but we can inspect the estimated mosquito mortality time series in the same spirit.
In particular, the weak periodic structure in the estimated $\epsilon_{dt}$ suggests that there may be some unmodeled, higher frequency process influencing transmission, either through the death rate itself or through another process.
Despite this, our estimated mosquito mortalities generally fell within the reasonable range from the literature \cite{Maciel-de-Freitas2008, Brady2013}.
The fact that our estimated mortality rate increases with temperature above 20 degrees also agrees with the empirical literature on mosquito survival \cite{Yang2009}, though our estimated relationship is generally more extreme than estimated in laboratory studies.

These observations suggest that our estimated mosquito mortality rates may actually be capturing the patterns of mosquito mortality in this system and are not obviously influenced by other sources of uncertainty or misspecification.
Further, this emphasizes that the dynamics of dengue fever, like many other infectious diseases \cite{Ellner1998,Koelle2004} and ecological systems \cite{Bjornstad2001}, are driven by the combined efforts of intrinsic non-linearities, seasonality, and stochasticity. 
Indeed, the driving force of these interacting processes is channelled largely through mosquito mortality.

\subsection*{Implications for control policy}

The importance of mosquito longevity in driving disease dynamics suggests the potential for control efforts that reduce the lifespan of adult mosquitoes to disrupt transmission by preventing mosquitoes from living long enough to progress through the extrinsic incubation period to the infectious state.
In fact, this forms the basis for much of the theory of adult control \cite{Burattini2008, Morrison2008, Smith2012}.
The fact that adult control (increasing the mortality rate) and larval control (decreasing the birth rate) produce nearly identical reductions in disease in our simulations suggests that the effect of control is likely driven by reducing the standing population of adult mosquitoes, rather than by limiting the transmission potential of exposed or infectious mosquitoes.

Discussion of the effectiveness of adult control often focuses on the fact that it produces only a brief decline in mosquito abundance \cite{Newton1992, Esu2010}.
When applied at the peak of an epidemic, this reduction in abundance is likely to have a relatively small effect, because the population rebounds while there are still a large number of infectious human hosts available \cite{Newton1992, Burattini2008}. 
However, our model demonstrates that reducing mosquito abundance early in the year can be highly effective at preventing dengue cases later in the year.
This effectiveness is driven primarily by the timing of the reduction in mosquito abundance and could be achieved by either adult or larval control strategies.
Although the mechanisms above might suggest further advantages for adult control, a mixed strategy of adult and larval control may be most effective in the long-term \cite{Burattini2008, Pinho2010}.
Fully exploring such combined strategies requires a more complete model of mosquito life history and the lags in population dynamics induced by the mosquito's aquatic stage.
These lags likely mean larval control should be applied prior to week 3 to achieve the desired reduction in the adult population in week 3.
Models of mosquito development can be very complex (e.g. \cite{Magori2009}) but our model could be extended to account for these lags through a compartment for the aquatic phase (as in \cite{Burattini2008, Pinho2010}), though at the cost of introducing several new parameters.

@@

It is important to note that the control we simulate is implemented over the top of the control policies already enacted by the city (and on which data are unavailable).
Also, the case data represents reports of "dengue-like illness", and as such could include cases of other diseases with similar symptoms (e.g. chikungunya).
Given the high underreporting rate expected for dengue fever, we expect unreported cases to far outnumber misreported cases.
Moreover, many of the pathogens that cause disease that might be mistaken for dengue are also vectored by \emph{Aedes aegypti}, and control policies driven by dengue fever are likely to aid in the control of other pathogens as well.

In addition, our model does not explicitly include all four serotypes.  
It is difficult to speculate about how important this might be.

The assumption of exponential wait times implicit in a compartmental model is perhaps ill-suited to modeling the race between mosquito lifespan and incubation period.
The long tails of the exponential distribution may overestimate the probability that a mosquito survives long enough to become infectious and underestimate the effect of decreasing mosquito lifespan.
More accurate models of the extrinsic incubation period could incorporate explicit delays (i.e., assuming a fixed incubation period \cite{Burattini2008}), or additional chained exposed compartments (i.e., assuming the incubation period is gamma-distributed, \cite{Lloyd2001}).
Including such mechanisms in our model would allow us to better differentiate the effects of larval and adult control and provide a more accurate picture of the effectiveness of adult control.
In fact, modeling studies that assume exponential wait times suggest that adult control is likely to be ineffective when applied during an epidemic \cite{Newton1992, Pinho2010}, while similar models that include explicit delays for the extrinsic incubation period show adult control being much more effective \cite{Burattini2008}.

>>>>

Given the central role that mosquito lifespan plays in driving dengue dynamics, we model fluctuations in the mosquito mortality rate using a flexible, stochastically forced harmonic oscillator.
The fact that this is the only source of stochasticity in the transmission process, however,  means that the estimated mosquito mortality time series could be soaking up other sources of stochasticity or model misspecification.
Our implementation of the mosquito mortality process is conceptually similar to the empirical forcing functions developed by Hooker and Ellner \cite{Hooker2015}.
In that framework, Hooker and Ellner \cite{Hooker2015} estimated nonparametric functions in time that modify either the differential equations or specific model parameters to provide a good fit to the data.
These forcing functions serve as residuals on the time derivatives, and can be more readily interpreted as indicators of lack-of-fit than residuals on the state variables.
We do not employ the same explicit goodness-of-fit testing framework as \cite{Hooker2015}, but we can inspect the estimated mosquito mortality time series in the same spirit.
In particular, the weak periodic structure in the estimated $\epsilon_{dt}$ suggests that there may be some unmodeled, higher frequency process influencing transmission, either through the mortality rate itself or through another process.

Despite this, our estimated mosquito mortalities generally fell within the reasonable range from the literature \cite{Maciel-de-Freitas2008, Brady2013}.
The fact that our estimated mortality rate increases with temperature above 20 degrees also agrees with the empirical literature on mosquito survival \cite{Yang2009}, though our estimated relationship is generally more extreme than estimated in laboratory studies, possibly as a result of assuming exponentially distributed extrinsic incubation period.
In particular, the assumption that the extrinsic incubation period is exponentially distributed, and thus that many mosquitoes experience a very brief incubation period, may overestimate the probability that a mosquito survives long enough to become infectious.
More accurate models of the extrinsic incubation period could incorporate explicit delays (i.e., assuming a fixed incubation period \cite{Burattini2008}), or additional chained exposed compartments (i.e., assuming the incubation period is gamma-distributed, \cite{Lloyd2001}).
These observations suggest that our estimated mosquito mortality rates may actually be capturing the patterns of mosquito mortality in this system and are not obviously influenced by other sources of uncertainty or misspecification.

\subsection*{Processes driving dynamics and implications for control}

Infectious diseases \cite{Ellner1998,Koelle2004} and ecological systems \cite{Bjornstad2001}, are often driven by the combined efforts of intrinsic non-linearities and feedbacks, seasonality, and stochasticity.
In this paper, we have highlighted the importance of these processes in driving the dynamics of dengue fever and shaping the relationship between mosquito abundance and human disease risk.

\subsubsection*{Adult control}

The importance of mosquito longevity in driving disease dynamics highlights the potential effectiveness of control efforts that target adult mosquitoes and disrupt transmission by preventing mosquitoes from living long enough to progress through the extrinsic incubation period to the infectious state.
In fact, this forms the basis for much of the theory of adult control \cite{Burattini2008, Morrison2008, Smith2012}.
Our results suggest that such control efforts (i.e., those that remove adult mosquitoes from the population) are most effective when applied late in the outbreak (week 24), after the number of cases has peaked.
If applied at the peak of the epidemic, the this control is likely less effective, as there are still a large number of infectious human hosts available \cite{Newton1992, Burattini2008}.
Control applied late in the outbreak pushes the system in the direction it is already heading.
Thus control that interacts with the existing seasonality is more effective.
In measles, this is also generally when the effect of stochasticity is most limited and the dynamics are pulled most strongly towards the deterministic attractor.
This is also generally when immunity is highest.

\subsubsection*{Larval controls}

In contrast, larval control is most effective early in the year, right at the beginning of the epidemic.
Intervention early in the year helps to prevent amplification and feedback.
Also often corresponds to the local peak in susceptibility.
This effectiveness is driven primarily by the timing of the reduction in mosquito abundance and could be achieved by either adult or larval control strategies.

The timing of the effectiveness of both of these control strategies is driven largely by the interplay of the intrinsic feedbacks and seasonality of the system, and in particular by the fluctuations in human immunity and mosquito abundance.
It is also interesting that neither strategy is most effective at the height of the outbreak, which highlights the fact that control strategies need to consider effects beyond the current season.

@@

It is important to note that the control we simulate is implemented over the top of the control policies already enacted by the city (and on which data are unavailable).

Also, the case data represents reports of "dengue-like illness", and as such could include cases of other diseases with similar symptoms (e.g. chikungunya).
Given the high underreporting rate expected for dengue fever, we expect unreported cases to far outnumber misreported cases.
Moreover, many of the pathogens that cause disease that might be mistaken for dengue are also vectored by \emph{Aedes aegypti}, and control policies driven by dengue fever are likely to aid in the control of other pathogens as well.
